---
title: "03_limma_pit"
output: md_document
---

DESeq2 is _not_ recommended for experiments with more than 100 samples ([see Mike Love's post](https://mikelove.wordpress.com/2016/09/28/deseq2-or-edger/)), so I decided to try the limma package. I followed [this tutorial](https://www.bioconductor.org/packages/devel/workflows/vignettes/RNAseq123/inst/doc/limmaWorkflow.html).

```{r}
library(tidyverse)
library(limma)
library(Glimma)
library(edgeR)
library(kableExtra)
library(cowplot)

# load custom functions
source("../R/functions.R")

knitr::opts_chunk$set(fig.path = '../figures/pit/',cache=TRUE)

```

First, I read in the data I processed in 00_datawrangling.Rmd.

```{r read}
# import "colData" which contains sample information and "countData" which contains read counts
colData <- read.csv("../results/00_colData_characterization.csv", header = T, row.names = 1)
countData <- read.csv("../results/00_countData_characterization.csv", header = T, row.names = 1)
geneinfo <- read.csv("../results/00_geneinfo.csv", row.names = 1)
```

```{r filter}
colData <- colData %>%
  dplyr::filter(grepl('pituitary', tissue)) %>%
  droplevels()
row.names(colData) <- colData$V1

# print sample sizes
colData %>% select(sex,treatment, tissue)  %>%  summary()

savecols <- as.character(colData$V1) 
savecols <- as.vector(savecols) 
countData <- countData %>% dplyr::select(one_of(savecols)) 

# check that row and col lenghts are equal
ncol(countData) == nrow(colData)
```

Then, I followed the steps from <https://github.com/macmanes-lab/RockDove/blob/master/parental_care/parental_analysis.Rmd>.

```{r edgeR}
# create a large DGEList with 3 elements
parentalobject <- DGEList(counts=countData, genes=geneinfo, group=colData$group)

# transform raw counts to countspermillion
cpms <- cpm(parentalobject)

# calculate number of lowly lowly expressed genes and remove them
table(rowSums(parentalobject$counts==0)==10)
keep_genes <- rowSums(cpms >= 1) >= 10
dge <- parentalobject[keep_genes, ]

# specific the design
parentaldesign <- model.matrix(~ colData$group )
colnames(parentaldesign) <- levels(colData$group)

# The TMM normalization
parentalobject <- calcNormFactors(parentalobject)
parentalobject <- estimateCommonDisp(parentalobject)
parentalobject <- estimateTagwiseDisp(parentalobject)
parentalobject <- estimateDisp(parentalobject, parentaldesign)
parentalobject <- estimateGLMCommonDisp(parentalobject, parentaldesign, verbose=TRUE)
parentalobject <- estimateGLMTrendedDisp(parentalobject, parentaldesign)
parentalobject <- estimateGLMTagwiseDisp(parentalobject, parentaldesign)

#  perform likelihood ratio test and thresholded testing
fit <- glmFit( parentalobject, parentaldesign, robust=T)
tr <- glmTreat(fit, lfc = 1)
topTags(tr)
```


# plotMDS (multidimential scaling)

```{r plotMDS}
plotMDS(parentalobject, cex = 0.5)
```

For color coding, I used this tutorial for guidance <https://combine-australia.github.io/RNAseq-R/06-rnaseq-day1.html>.

```{r plotMDS-lables}
levels(colData$treatment)
col.treatment <- c("#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c", "#fdbf6f", "#ff7f00", "#cab2d6")[colData$treatment]

plotMDS(parentalobject,col=col.treatment, labels = colData$sex)
legend("bottomleft",fill=c("#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c", "#fdbf6f", "#ff7f00", "#cab2d6"),legend=levels(colData$treatment))
title("Pituitary Colored by Treatment")

plotMDS(parentalobject,dim=c(3,4), col=col.treatment, labels = colData$sex)
legend("bottomright",fill=c("#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c", "#fdbf6f", "#ff7f00", "#cab2d6"),legend=levels(colData$treatment))
title("Pituitary Colored by Treatment")
```

# specify contrasts and make MA plots 

```{r 01-contrasts}
# view all levels
levels(colData$group)

# subset of conrasts - sex specific comparing hatch to lay
my.contrasts <- makeContrasts(
	         FP_CB = female.pituitary.control - female.pituitary.bldg,
	         FP_BL = female.pituitary.bldg - female.pituitary.lay,
	         FP_Li3 = female.pituitary.lay - female.pituitary.inc.d3,
	         FP_i39 = female.pituitary.inc.d3 - female.pituitary.inc.d9,
	         FP_i917 = female.pituitary.inc.d9 - female.pituitary.inc.d17,
	         FP_i17H = female.pituitary.inc.d17 - female.pituitary.hatch,
	         FP_H5 = female.pituitary.hatch -  female.pituitary.n5,
	         FP_n59 = female.pituitary.n5 - female.pituitary.n9,
	         FP_n9C = female.pituitary.n9 - female.pituitary.control,
	         
	         MP_CB = male.pituitary.control - male.pituitary.bldg,
	         MP_BL = male.pituitary.bldg - male.pituitary.lay,
	         MP_Li3 = male.pituitary.lay - male.pituitary.inc.d3,
	         MP_i39 = male.pituitary.inc.d3 - male.pituitary.inc.d9,
	         MP_i917 = male.pituitary.inc.d9 - male.pituitary.inc.d17,
	         MP_i17H = male.pituitary.inc.d17 - male.pituitary.hatch,
	         MP_H5 = male.pituitary.hatch -  male.pituitary.n5,
	         MP_n59 = male.pituitary.n5 - male.pituitary.n9,
	         MP_n9C = male.pituitary.n9 - male.pituitary.control,
	         
	         FP_n9B = female.pituitary.n9 - female.pituitary.bldg,
	         MP_n9B = male.pituitary.n9 - male.pituitary.bldg,
levels=parentaldesign)

# create a list with all the two way contrasts

mycontrasts <- c("FP_CB", "FP_BL", "FP_Li3", "FP_i39", "FP_i917", "FP_i17H", "FP_H5", "FP_n59", "FP_n9C",
                 "MP_CB", "MP_BL", "MP_Li3", "MP_i39", "MP_i917", "MP_i17H", "MP_H5", "MP_n59", "MP_n9C",
                 "FP_n9B", "MP_n9B")

# use the printplotcontrasts function to print summary stats and a volcano plot

for(i in mycontrasts){
  printplotcontrasts(i)
}

```

# volcano plots

```{r volcanoplots}
# from http://www.compbio.dundee.ac.uk/user/pschofield/Projects/teaching_pg/workshops/biocDGE.html#maplots

lrt <- glmLRT(fit,coef=2)
topTags(lrt)

tt <- topTags(lrt,n=10000)$table

ggplot(data=tt) + geom_point(aes(x=logFC,y=-log(FDR),color=logCPM)) +
  scale_colour_gradientn(colours=c("#000000" ,"#FF0000" ))
```